Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-09-05 paper/main.tex
format supplement subsection headers
adjust some formulations

Mechanical properties of rod-shaped bacteria have been studied
extensively~\cite{Chatterjee1988,Takeuchi2005,IWAI2002}.
It was found that the MreB protein is the driving force in maintaining their distinct elongated
shape~\cite{Ursell2014}, which plays an important role in the dynamics of
growth~\cite{Billaudeau2017}, division~\cite{Harry2001} and their collective
behaviour~\cite{vanGestel2015}.
Furthermore, it has been discovered that the rod can be bent thus introducing temporary elastic or
permanent plastic deformations~\cite{Amir2014_2}.
They are an interesting target system since their physical form provides more surface information
compared to spherical cells.
In addition, parameters such as growth rates have been found to vary across multiple
cells~\cite{Koutsoumanis2013} which highlights the need to incorporate this heterogeneity on a
single-cell level.

Computational models for rod-shaped bacteria allow researchers to systematically analyze their
properties and infer emerging phenomena~\cite{Cho2007,Jnsson2005}.
Existing tools can describe growth, division and many effects such as colony formation and
expansion~\cite{Matyjaszkiewicz2017,Doumic2020} but they are limited in describing the
aforementioned plastic or elastic deformations of the rod.
Furthermore, the literature on how to estimate the parameters of such models is very sparse.

@@

%###################################################################################################
\section{Overview}
This work consists of many subcomponents which can be combined in order to achieve various tasks.
Figure~\ref{fig:flowchart-project-structure} shows a high-level overview of all components.
We assume that bacterial rods can be represented by a colleciton of vertices $\vec{x}_i$.
This assumption alone is enough to enable a cascade of other components.
Using the spatial representation, we provide visualization methods that create 3D models from the
given vertices which can then either be rendered or projected to a 2D plane.
The data extraction component determines the vertices $\vec{x}_i$ from given cell masks which
correspond to microscopic images.
This allows us in particular to turn any sequence of images (movie) into a sequence of bacterial
positions.
The dynamics of the vertices $\vec{x}_i$ are determined by a computational model which is specific
to the respective use-case and should be defined by the researcher.
It merely needs to utilize the aforementioned spatial representation of bacterial rods as a
collection of vertices $\vec{x}_i$.
This work will provide one example of how such a model can be defined.
Notably, the visualization and data extraction components do not require any knowledge about the
computational model, thus allowing us to reuse their functionality across multiple models.

@@

The mechanical behaviour of rod-shaped bacteria is governed by a plethora of effects.
In this work, we aim to establish a model which is abstract enough such that it can be reused for
many types of rod-shaped bacteria but also explicit enough such that we can quantify its parameters.
We are thus aiming to work with a set of assumptions that is as general as possible but still
specific to rod-shaped bacteria.
With this delicate balance in mind, it is clear that many cellular processes which might be specific
to one particular celltype or simply not necessary to assume in order to construct our model, will
be omitted alltogether.
This concerns effects such as polar interactions~\cite{Duvernoy2018}, extracellular
reactions~\cite{Wang2010} or differentiation processes~\cite{vanGestel2015,Lpez2010} and many other.
We restrict ourselves to the aspects presented in table~\ref{table:simulation-aspects}.
For a collection of other known effects which could be considered, we refer to the supplementary
section~\ref{section:discussion}, and~\ref{section:supplement-more-simulation-aspects}.

@@

We implemented the described model in \texttt{cellular\_raza}~\cite{Pleyer2025}.
Doing so has many benefits.
Not only, was a more simpler model already available which could be adapted to our needs.
But furthermore, due to the nature of \texttt{cellular\_raza} in that it does not assume any
particular cellular representation, we are free to change the underlying model at any time, even
making fundamental changes.
It is an \ac{abm} which is able to natively track cells through the course of a numerical simulation
and also keeps track of parent/daughter relationships.
This allows us to to experiment with different model implementations and still retain performance.
By leveraging the strong safety guarantees provided by the Rust programming language, we eliminate a
whole class of problems from the start.
% TODO CITATION
In addition, we are using the \texttt{pyo3} and \texttt{maturin} crates which allow us to easily
generate python bindings and thus interoperate between the two ecosystems.
While this particular model is more complex due to the nature of non-trivial (asymmetric)
interactions, we were still able to properly utilize the speedup from executing simulations in
parallalel using the built-in parallelization methods of \texttt{cellular\_raza} (see
Supplementary Material, Figure~\ref{fig:performance}).

@@

For visualization we use pyvista, which provides high-level wrappers around the popular VTK
library~\cite{vtkBook,Sullivan2019}.
The first stage of our visualization is to generate 3D meshes for each bacterium, which we consider
an intermediate representation.
We insert spheres of a given radius $r$ at each vertex $\vec{x}_i$ and connect adjacent vertices
$\vec{x}_i,\vec{x}_{i+1}$ with a cylinder, also of radius $r$ which is positioned such that its core
is identical to the connecting line segment between the two vertices.
Finally, we combine the meshes of all of these objects to obtain a single mesh of one bacterium.
A render of this result can be seen in figure~\ref{fig:progression-image-generation}.
Since we use a 3D rendering engine as an intermediate representation, the resulting image can have
overlaps of bacteria which are placed at higher or lower values along the z-direction.

To generate masks, we assign unique colors to every mesh of the respective cellular agent.
This mapping is unique, ensuring no two agents will use identical colors.
We then use a parallel projection and remove any existing light sources in order to obtain a solid
color for every cell-agent.
The resulting image is shown in subfigure (B).
Subfigure (C) shows an example of a simplified way to generate microscopic images.
The properties of the meshes such as roughness and reflectivity have been adjusted and we use
physically-based rendering of VTK.
The result of this render is augmented by introducing white noise on the pixel values at each point.
This approach is very rudimentary and we hope to extend these methods in the future.
In order to properly generate data, we need to deal with microscopic imaging techniques and possible
defects that can occurr.
It is crucial to get this part of the image-generation process right in order to obtain
near-realistic images that can be further reused.

@@

In order to determine if our extraction algorithm provides reliable results, we have to take
advantage of a functionality which will be introduced later in
section~\ref{subsection:visualization}.
Our methodology makes use of the fact that we have already set up our model and solving scheme and
are able to produce cell masks from these numerical results.
In doing so, we have exact knowledge about the position of our agents since we can simply obtain
this information from our model.
But on the other hand we can also try to extract this position from the generated cell-masks.
By comparing the differences between them, we get a good estimate of how well our extraction
algorithm can determine the true position of the bacteria.

@@

\appendix
%###################################################################################################
\section*{Supplementary Material}

\renewcommand{\thesection}{S\arabic{section}}

%###################################################################################################
\section{Data Preparation}
\begin{enumerate}
    \item Have masks of data; reference image segmantation tools; or do it by hand
    \item ensure that colors are matching for bacteria between time-steps
    \item clean borders of images if needed
    \item know domain size; otherwise use pixels
\end{enumerate}

%###################################################################################################
\section{Parameter Estimation}
\label{sec:supplement-parameter-profiles}
%TODO {explain profiles}
\begin{figure}
    \centering
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-0.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-1.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-2.pdf}\\%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-3.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-4.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-5.pdf}
    \caption{Parameter estimation of growth rates for individual cells}%TODO extend caption}
    \label{fig:parameter-estimates-supplement-growth-rates}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-0.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-1.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-2.pdf}\\%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-3.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-4.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-5.pdf}%
    \caption{Parameter estimation of radii for individual cells}%TODO extend caption}
    \label{fig:parameter-estimates-supplement-radii}
\end{figure}

%###################################################################################################
\section{More Simulation Aspects}
\label{section:supplement-more-simulation-aspects}
%TODO insert this into the discussion later
\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material either along the circular part of the
            rod or at the tip~\cite{Robert2014,Takeuchi2005}.\\
        (3) & Differentiation &
            \textit{B.subtilis} is known to differentiate into matrix-producing and
            surfactin-producing cells \cite{vanGestel2015,Lpez2010}.\\
        (4) & Division &
            The formation and continuation of van Gogh bundles is driven by cell division
            \cite{vanGestel2015}.
            Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        (5) & Variable Parameters &
            Parameters for individual cells are not fixed values but rather taken from a
            distribution \cite{Koutsoumanis2013}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (6) & Adhesion &
            Bacteria adhere to each other at longer distances and attach when in close contact
            \cite{Verwey1947,Trejo2013}.
            The interaction between rods can be polarized \cite{Duvernoy2018}.\\
        (7) & Friction &
            Friction between cells \cite{Grant2014} can be asymmetrical \cite{Doumic2020}.\\
        &\textbf{(DC) Domain-Cell Interactions}\\
        \midrule
        (8) & External Forces &
            Bacteria tend to stick to surfaces \cite{vanLoosdrecht1989}.
            The extracellular gel exerts a force onto the cells \cite{Grant2014}.\\
        (9) & Extracellular Reactions &
            Bacteria can take up nutrients or possible secrete/take up signalling molecules
            \cite{Li2025}.\\
        \bottomrule
    \end{tabularx}
    \label{table:simulation-aspects-supplement}
    \caption{
        This table extends table~\ref{table:simulation-aspects} with additional observations.
        It has to be noted that this table includes more specific behaviour that might not be
        generalizable across all types of rod-shaped bacteria.
    }
\end{table}

%###################################################################################################
\section{Performance}

@@

%###################################################################################################
\section{Multilayer Predictions}
% TODO

>>>>

Mechanical properties of rod-shaped bacteria have been studied in great detail
experimentally~\cite{Chatterjee1988,Takeuchi2005,IWAI2002}.
It was found that the MreB protein is the driving force in maintaining their distinct elongated
shape~\cite{Ursell2014}, which plays an important role in the dynamics of
growth~\cite{Billaudeau2017}, division~\cite{Harry2001} and their collective
behaviour~\cite{vanGestel2015}.
Furthermore, it has been discovered that the rod can be bent thus introducing temporary elastic or
permanent plastic deformations~\cite{Amir2014_2}.
They are an interesting target system since their physical form provides more surface information
compared to spherical cells.
In addition, parameters such as growth rates have been found to vary across multiple
cells~\cite{Koutsoumanis2013} which highlights the need to incorporate this heterogeneity on a
single-cell level.

Computational models for rod-shaped bacteria allow researchers to systematically analyze their
properties and infer emerging phenomena~\cite{Cho2007,Jnsson2005}.
Existing tools can describe growth, division and many effects such as colony formation and
expansion~\cite{Matyjaszkiewicz2017,Doumic2020} but they are limited in describing the
aforementioned plastic or elastic deformations of the rod.
Furthermore, the literature on how to estimate the parameters of such models directly from data is
very sparse.

@@

%###################################################################################################
\section{Overview}
This work consists of many subcomponents which can be combined in order to achieve various tasks.
Figure~\ref{fig:flowchart-project-structure} shows a high-level overview of all components.
We assume that bacterial rods can be represented by a colleciton of vertices $\vec{x}_i$ (see
section~\ref{subsec:spatial-representation}).
This assumption alone is enough to enable a cascade of other components.
Using the spatial representation, we provide visualization methods that create 3D models from the
given vertices which can then either be rendered or projected to a 2D plane.
The data extraction component determines the vertices $\vec{x}_i$ from given cell masks which
correspond to microscopic images.
This allows us in particular to turn any sequence of images (movie) into a sequence of bacterial
positions.
The dynamics of the vertices $\vec{x}_i$ are determined by a computational model which is specific
to the respective use-case and should be defined by the researcher.
It merely needs to utilize the aforementioned spatial representation of bacterial rods as a
collection of vertices $\vec{x}_i$.
This work will provide one example of how such a model can be defined.
Notably, the visualization and data extraction components do not require any knowledge about the
computational model, thus allowing us to reuse their functionality across multiple models.

@@

The mechanical behaviour of rod-shaped bacteria is governed by a plethora of effects.
In this work, we aim to establish a model which is abstract enough such that it can be reused for
many types of rod-shaped bacteria but also explicit enough such that we can quantify its parameters.
We are thus aiming to work with a set of assumptions that is as general as possible but still
specific to rod-shaped bacteria.
With this delicate balance in mind, it is clear that many cellular processes which might be specific
to one particular celltype or simply not necessary to assume in order to construct our model, will
be omitted alltogether.
This concerns effects such as polar interactions~\cite{Duvernoy2018}, extracellular
reactions~\cite{Wang2010} or differentiation processes~\cite{vanGestel2015,Lpez2010} and many other.
When describing our system, we distinguish between cellular aspects which ocurr within a single
cell and cell-cell interactions between numerous bacteria.
We restrict ourselves to the aspects presented in table~\ref{table:simulation-aspects}.
For a collection of other known effects which could be considered, we refer to the supplementary
section~\ref{section:discussion}, and~\ref{section:supplement-more-simulation-aspects}.

@@

We implemented the described model in \texttt{cellular\_raza}~\cite{Pleyer2025}.
Doing so has many benefits.
Not only, was a more simpler model already available which we were able to adapt to our needs.
But by design \texttt{cellular\_raza} aims to provide researchers with a great amount of flexibility
and therefore does not assume any particular cellular representation, thus enabling us to make
fundamental changes.
This allows us to to experiment with different model implementations and still retain performance.
Furthermore, it is an \ac{abm} which is able to natively track cells through the course of a
numerical simulation and also keeps track of mother/daughter relationships.
On the programming side, we are using the \texttt{pyo3} and \texttt{maturin} crates which allow us
to easily generate Python bindings and enable us to interoperate between the two ecosystems.
By using \texttt{cellular\_raza}, we can make use of multiple threads for running a single numerical
simulation, using the built-in parallelization methods (see Supplementary Material,
Figure~\ref{fig:performance}).

@@

To visualize the spatial configuration of our results we use pyvista, which provides high-level
wrappers around the popular VTK library~\cite{vtkBook,Sullivan2019}.
The first stage of our visualization is to generate 3D meshes for each bacterium, which we consider
an intermediate representation.
We insert spheres of a given radius $r$ at each vertex $\vec{x}_i$ and connect adjacent vertices
$\vec{x}_i,\vec{x}_{i+1}$ with a cylinder, also of radius $r$ which is positioned such that its core
is identical to the connecting line segment between the two vertices.
Finally, we combine the meshes of all of these objects to obtain a single mesh of one bacterium.
A render of this result can be seen in figure~\ref{fig:progression-image-generation}.
Since we use a 3D rendering engine as an intermediate representation, the resulting image can have
overlaps of bacteria which are placed at higher or lower values along the z-direction.

To generate masks, we assign unique colors to every mesh of the respective cellular agent, ensuring
no two agents will use identical colors.
We then use a parallel projection and remove any existing light sources in order to obtain a solid
color for every cell-agent.
The resulting image is shown in subfigure (B).
Subfigure (C) shows an example of a simplified way to generate microscopic images.
The properties of the meshes such as roughness and reflectivity have been adjusted and we use
physically-based rendering of VTK.
The result of this render is augmented by introducing white noise on the pixel values at each point.
This approach is very rudimentary and we hope to extend these methods in the future.
In order to properly generate data, we need to deal with microscopic imaging techniques and possible
defects that can occurr.
It is crucial to get this part of the image-generation process right in order to obtain
near-realistic images that can be further reused.

@@

In order to determine if our extraction algorithm provides reliable results, we have to take
advantage of a functionality which will be introduced later in
section~\ref{subsection:visualization}.
Our methodology makes use of the fact that we have already set up our model and solving scheme and
are able to produce cell masks from these numerical results.
In doing so, we have exact knowledge about the position of our agents since we can simply obtain
this information from our computational model.
But on the other hand we can also try to extract these positions from the generated cell-masks.
By comparing the differences between them, we get a good estimate of how well our extraction
algorithm can determine the true position of the bacteria.

@@

\appendix
%###################################################################################################
\renewcommand{\thesection}{}
\renewcommand{\thesubsection}{S\arabic{subsection}}
\section{Supplementary Material}

%###################################################################################################
\subsection{Data Preparation}
\begin{enumerate}
    \item Have masks of data; reference image segmantation tools; or do it by hand
    \item ensure that colors are matching for bacteria between time-steps
    \item clean borders of images if needed
    \item know domain size; otherwise use pixels
\end{enumerate}

%###################################################################################################
\subsection{Parameter Estimation}
\label{sec:supplement-parameter-profiles}
%TODO {explain profiles}
\begin{figure}
    \centering
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-0.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-1.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-2.pdf}\\%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-3.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-4.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/growth-rate-5.pdf}
    \caption{Parameter estimation of growth rates for individual cells}%TODO extend caption}
    \label{fig:parameter-estimates-supplement-growth-rates}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-0.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-1.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-2.pdf}\\%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-3.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-4.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_fit/profiles/radius-5.pdf}%
    \caption{Parameter estimation of radii for individual cells}%TODO extend caption}
    \label{fig:parameter-estimates-supplement-radii}
\end{figure}

%###################################################################################################
\subsection{More Simulation Aspects}
\label{section:supplement-more-simulation-aspects}
%TODO insert this into the discussion later
\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material either along the circular part of the
            rod or at the tip~\cite{Robert2014,Takeuchi2005}.\\
        (3) & Differentiation &
            \textit{B.subtilis} is known to differentiate into matrix-producing and
            surfactin-producing cells \cite{vanGestel2015,Lpez2010}.\\
        (4) & Division &
            The formation and continuation of van Gogh bundles is driven by cell division
            \cite{vanGestel2015}.
            Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        (5) & Variable Parameters &
            Parameters for individual cells are not fixed values but rather taken from a
            distribution \cite{Koutsoumanis2013}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (6) & Adhesion &
            Bacteria adhere to each other at longer distances and attach when in close contact
            \cite{Verwey1947,Trejo2013}.
            The interaction between rods can be polarized \cite{Duvernoy2018}.\\
        (7) & Friction &
            Friction between cells \cite{Grant2014} can be asymmetrical \cite{Doumic2020}.\\
        &\textbf{(DC) Domain-Cell Interactions}\\
        \midrule
        (8) & External Forces &
            Bacteria tend to stick to surfaces \cite{vanLoosdrecht1989}.
            The extracellular gel exerts a force onto the cells \cite{Grant2014}.\\
        (9) & Extracellular Reactions &
            Bacteria can take up nutrients or possible secrete/take up signalling molecules
            \cite{Li2025}.\\
        \bottomrule
    \end{tabularx}
    \label{table:simulation-aspects-supplement}
    \caption{
        This table extends table~\ref{table:simulation-aspects} with additional observations.
        It has to be noted that this table includes more specific behaviour that might not be
        generalizable across all types of rod-shaped bacteria.
    }
\end{table}

%###################################################################################################
\subsection{Performance}

@@

%###################################################################################################
\subsection{Multilayer Predictions}
% TODO
